---
title: Calculation of Reef Health Indicators - V6
subtitle: Prepared by UB-ERI for HRHP
output:
  html_document:
    fig_height: 6
    fig_width: 8
    collapse: no
    code_folding: show
    theme: sandstone
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
    code_download: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r start1, message=FALSE, warning=FALSE, results='hide'}
# Load in packages
library(readxl) # read excel files
library(tidyverse) # manipulate, tidy data, make plots
library(knitr) # increase markdown capabilities
```


This script follows the documentation outlined in the following document:
Calculation of Indicator Values ([link](https://docs.google.com/document/d/14GapQRfsUaCF2kV6RtnJ_H1pB5CvKAUF/edit?usp=sharing&ouid=104670669775847416587&rtpof=true&sd=true)). This script runs using the AGRRA standard format for data (as of June 2024). Other formats may be used as long as data are prepared accordingly, as described in the documentation.


# Percent Cover Indicator Calculations

### 1 - Data

```{r benthicdata, message=FALSE, warning=FALSE, results='hide'}
# Metadata
df_metadata_survey_V0 <- read_excel("Metadata.xlsx", sheet = 3, trim_ws = TRUE, na = "")
df_metadata_benthiccovercategory_V0 <- read_excel("Metadata.xlsx", sheet = 5, trim_ws = TRUE, na = "")
df_metadata_benthiccover_V0 <- read_excel("Metadata.xlsx", sheet = 6, trim_ws = TRUE, na = "")

# Benthic
df_benthic_transect_V0 <- read_excel("BenthicRaw.xlsx", sheet = 3, trim_ws = TRUE, skip = 1, na = "")
df_benthic_cover_raw <- read_excel("BenthicRaw.xlsx", sheet = 4, skip = 1, na = "", col_types = c("numeric", "numeric", "text", "numeric", "numeric", "text", "numeric"))
```


### 2 - Data Preparation

```{r benthic1, fig.height=8, fig.width=14}
# Rename ID columns for appropriate merging
df_benthic_transect <- df_benthic_transect_V0 %>%
  rename(Transect = ID)
df_metadata_benthiccover_primary <- df_metadata_benthiccover_V0 %>%
  rename(Primary = ID, Primary_Category = Category) # primary observations
df_metadata_benthiccover_secondary <- df_metadata_benthiccover_V0 %>%
  rename(Secondary = ID, Secondary_Category = Category) # secondary observations
df_metadata_benthiccovercategory_primary <- df_metadata_benthiccovercategory_V0 %>%
  rename(Primary_Category = ID) # primary observations
df_metadata_benthiccovercategory_secondary <- df_metadata_benthiccovercategory_V0 %>%
  rename(Secondary_Category = ID) # secondary observations
df_metadata_survey <- df_metadata_survey_V0 %>%
  rename(Survey = ID)

# Using common keys, left-join dataframes to Cover Raw, and rename duplicate headers for clarity
benthic_cover_merged_V0 <- df_benthic_cover_raw %>%
  left_join(df_benthic_transect, by = "Transect") %>%
  left_join(df_metadata_benthiccover_primary, by = "Primary") %>%
  left_join(df_metadata_benthiccover_secondary, by = "Secondary") %>%
  left_join(df_metadata_benthiccovercategory_primary, by = "Primary_Category") %>%
  left_join(df_metadata_benthiccovercategory_secondary, by = "Secondary_Category") %>%
  left_join(df_metadata_survey, by = "Survey") %>%
  rename(Primary_Category_Name = Name.x.x, Secondary_Category_Name = Name.y.y, Site = Name, Primary_State = State.x, Secondary_State = State.y)

# Create column counting the total observed points for a transect, Ntran
Ntran_data <- benthic_cover_merged_V0 %>%
  group_by(Subregion, Transect, Site) %>%
  summarize(Ntran = n())

# Add Ntran data to benthic cover df
benthic_cover_merged_V1 <- benthic_cover_merged_V0 %>%
  left_join(Ntran_data, by = c("Subregion", "Transect", "Site")) %>%
  # change category of newly dead  so as not to count newly dead in percent cover
  mutate(
    Primary_Category_Name = case_when(
      Primary_State == "Newly Dead" ~ "(newly dead)",
      TRUE ~ Primary_Category_Name
    ),
    Secondary_Category_Name = case_when(
      Secondary_State == "Newly Dead" ~ "(newly dead)",
      TRUE ~ Secondary_Category_Name
    )
  )

# Define function to assign coral weight based on primary and secondary observations
calculate_coral_cover <- function(primary, secondary) {
  if (!is.na(primary) && !is.na(secondary) && primary == "Calcifiers :: Coral" && secondary == "Calcifiers :: Coral") {
    return(1) # 100% coral if primary and secondary are real, and both are coral
  } else if (!is.na(primary) && primary == "Calcifiers :: Coral" && is.na(secondary)) {
    return(1) # ELSE 100% coral if primary real, primary coral, and secondary NA
  } else if ((!is.na(primary) && primary == "Calcifiers :: Coral") || (!is.na(secondary) && secondary == "Calcifiers :: Coral")) {
    return(0.5) # ELSE 50% coral if primary real and coral, or secondary real and coral
  } else {
    return(0) # ELSE 0% coral
  }
}

# Define function to assign algae weight based on primary and secondary observations
calculate_algae_cover <- function(primary, secondary) {
  if (!is.na(primary) && !is.na(secondary) && primary == "Algae :: Macro :: Fleshy" && secondary == "Algae :: Macro :: Fleshy") {
    return(1) # 100% algae if primary and secondary are real, and both are algae
  } else if (!is.na(primary) && primary == "Algae :: Macro :: Fleshy" && is.na(secondary)) {
    return(1) # ELSE 100% algae if primary real, primary algae, and secondary NA
  } else if ((!is.na(primary) && primary == "Algae :: Macro :: Fleshy") || (!is.na(secondary) && secondary == "Algae :: Macro :: Fleshy")) {
    return(0.5) # ELSE 50% algae if primary real and algae, or secondary real and algae
  } else {
    return(0) # ELSE 0% algae
  }
}

# Add column for coral and algae cover weight (0, 0.5, 1) based on Primary and Secondary organisms using the new functions
benthic_cover_merged_V2 <- benthic_cover_merged_V1 %>%
  mutate(Coral_Cover_Weight = mapply(calculate_coral_cover, Primary_Category_Name, Secondary_Category_Name)) %>%
  mutate(Algae_Cover_Weight = mapply(calculate_algae_cover, Primary_Category_Name, Secondary_Category_Name)) %>%
  # Select only relevant columns for simplicity
  select(
    Subregion, Transect, Primary_Category_Name, Secondary_Category_Name,
    Coral_Cover_Weight, Algae_Cover_Weight, Site, Code, Ntran
  )
```

### 3. Calculations

```{r benthic3, fig.height=8, fig.width=14}
# Calculate transect-level percent cover as sum of coral/algae weight for each point, divided by total points on transect
benthic_cover_summarized <- benthic_cover_merged_V2 %>%
  group_by(Subregion, Transect, Site, Code, Ntran) %>%
  summarize(
    Coral_Cover = 100 * sum(Coral_Cover_Weight) / Ntran,
    Algae_Cover = 100 * sum(Algae_Cover_Weight) / Ntran
  )
```

```{r benthic4, fig.height=8, fig.width=14}
# Calculate percent cover average for sites as mean coral/algae transect cover for each site
benthic_cover_site_summarized <- benthic_cover_summarized %>%
  group_by(Subregion, Site, Code) %>%
  summarise(
    Mean_Coral_Cover = mean(Coral_Cover),
    Mean_Algae_Cover = mean(Algae_Cover)
  )
```

```{r benthic5, fig.height=8, fig.width=14}
# Calculate percent cover average for subregions as mean coral/algae mean site cover for each subregion
benthic_cover_subregion_summarized <- benthic_cover_site_summarized %>%
  group_by(Subregion) %>%
  summarise(
    Mean_Coral_Cover = mean(Mean_Coral_Cover),
    Mean_Algae_Cover = mean(Mean_Algae_Cover)
  )
```



### 4. Comparing the Results

The following script provides a way to compare these results to HRI official results. To change subregion, adjust the variable subregion_of_interest.

```{r benthiccompare, fig.height=8, fig.width=14}
# Load in RHI 2024 calculated values
HRI_RHI <- read_excel("RHI 2024 Redo calculations with new Turneffe for reviewApril 9 2024.xlsx",
  sheet = 1, trim_ws = TRUE, na = ""
)

# Select which subregion you will compare to the HRI results
subregion_of_interest <- "Central Barrier Complex"

# Restructure the data to be comparable
HRI_RHI_subregion <- HRI_RHI %>%
  filter(Subregion == subregion_of_interest) %>%
  mutate(Mean_Coral_Cover_HRI = `Live Coral (%)` * 100) %>%
  mutate(Mean_Algae_Cover_HRI = `Fleshy Macroalgae (%)` * 100) %>%
  mutate(Code_Name = paste(Code, Name)) %>%
  select(Code_Name, Mean_Coral_Cover_HRI, Mean_Algae_Cover_HRI)
calculated_RHI_subregion <- benthic_cover_site_summarized %>%
  ungroup() %>%
  filter(Subregion == subregion_of_interest) %>%
  mutate(
    Code_Name = paste(Code, Site),
    Mean_Coral_Cover_Calculated = Mean_Coral_Cover,
    Mean_Algae_Cover_Calculated = Mean_Algae_Cover
  ) %>%
  select(Code_Name, Mean_Coral_Cover_Calculated, Mean_Algae_Cover_Calculated)

# Create dataset with both HRI and calculated values
compare_benthic <- HRI_RHI_subregion %>%
  left_join(calculated_RHI_subregion, by = "Code_Name") %>%
  filter(Code_Name != "NA NA")
```

```{r benthiccompare_coral, fig.height=8, fig.width=14}
# Compare coral
compare_benthic_coral <- compare_benthic %>%
  select(Code_Name, Mean_Coral_Cover_HRI, Mean_Coral_Cover_Calculated) %>%
  mutate(Match = case_when(
    abs(Mean_Coral_Cover_HRI - Mean_Coral_Cover_Calculated) < 1e-3 ~ "Yes", # allow for tiny differences
    TRUE ~ "No"
  ))
compare_benthic_coral
```

```{r benthiccompare_algae, fig.height=8, fig.width=14}
# Compare algae
compare_benthic_algae <- compare_benthic %>%
  select(Code_Name, Mean_Algae_Cover_HRI, Mean_Algae_Cover_Calculated) %>%
  mutate(Match = case_when(
    abs(Mean_Algae_Cover_HRI - Mean_Algae_Cover_Calculated) < 1e-3 ~ "Yes", # allow for tiny differences
    TRUE ~ "No"
  ))
compare_benthic_algae
```



# Fish Biomass Indicator Calculations

### 1 - Data

```{r start2fish, message=FALSE, warning=FALSE, results='hide'}
# Metadata
df_metadata_survey_V0 <- read_excel("Metadata.xlsx", sheet = 3, trim_ws = TRUE, na = "")
df_metadata_fishtaxonomy_V0 <- read_excel("Metadata.xlsx", sheet = 8, trim_ws = TRUE, skip = 1, na = "")

# Fish
df_fish_transect_V0 <- read_excel("FishRaw.xlsx", sheet = 3, trim_ws = TRUE, skip = 1, na = "")
df_fish_counts <- read_excel("FishRaw.xlsx", sheet = 4, trim_ws = TRUE, skip = 0, na = "")

# Additional
df_biomass_calcs <- read.csv("fish_biomass_specification.csv", na.strings = "")
```

### 2 - Creating a Merged Dataframe

```{r fish1, fig.height=8, fig.width=14}
# Rename ID columns
df_metadata_survey <- df_metadata_survey_V0 %>%
  rename(Survey = ID)
df_fish_transect <- df_fish_transect_V0 %>%
  rename(Transect = ID)
df_metadata_fishtaxonomy <- df_metadata_fishtaxonomy_V0 %>%
  rename(Taxonomy = ID)
# Clean biomass reference data
df_biomass_calcs_clean <- df_biomass_calcs %>%
  mutate(
    Family = word(Name, 1),
    Binomial = ifelse(str_detect(Name, " "), word(Name, 2, 3), NA_character_)
  ) %>%
  separate(Binomial, into = c("Genus", "Species"), sep = " ", fill = "right")

# Merge datasets together
merged_fish <- df_fish_counts %>%
  left_join(df_fish_transect, by = "Transect") %>%
  left_join(df_metadata_survey, by = "Survey") %>%
  left_join(df_metadata_fishtaxonomy, by = "Taxonomy") %>%
  left_join(df_biomass_calcs_clean, by = c("Family", "Genus", "Species"))
```

### 3 - Data Preparation

```{r fish2, fig.height=8, fig.width=14}
# Specify biomass category of fish families
merged_fish_targets <- merged_fish %>%
  filter(LWRconv == 1) %>%
  mutate(Biomass_Category = case_when(
    Family == "Acanthuridae" ~ "H",
    Family == "Scaridae" ~ "H",
    Family == "Epinephelidae" ~ "C",
    Family == "Lutjanidae" ~ "C"
  )) %>%
  # assign placeholder values for size class
  mutate(Size = case_when(
    `Size Class` == "0 - 5cm" ~ 2.5,
    `Size Class` == "6 - 10cm" ~ 8,
    `Size Class` == "11 - 20cm" ~ 15.5,
    `Size Class` == "21 - 30cm" ~ 25.5,
    `Size Class` == "31 - 40cm" ~ 35.5,
    `Size Class` == "50cm" ~ 50,
    `Size Class` == "60cm" ~ 60,
    `Size Class` == "70cm" ~ 70,
    `Size Class` == "80cm" ~ 80,
    `Size Class` == "90cm" ~ 90,
    `Size Class` == "100cm" ~ 100,
    `Size Class` == "110cm" ~ 110,
    `Size Class` == "120cm" ~ 120,
    `Size Class` == "130cm" ~ 130,
    `Size Class` == "140cm" ~ 140,
    `Size Class` == "150cm" ~ 150,
    `Size Class` == "160cm" ~ 160,
    `Size Class` == "170cm" ~ 170,
    `Size Class` == "180cm" ~ 180,
    `Size Class` == "190cm" ~ 190,
    `Size Class` == "200cm+" ~ 200
  ))
```

### 4 - Calculations

```{r fish3, fig.height=8, fig.width=14}
# Ensure all combinations of transect and biomass category are established
complete_grid_fish <- expand_grid(
  Transect = unique(merged_fish_targets$Transect),
  Biomass_Category = unique(merged_fish_targets$Biomass_Category)
)

# Calculate biomass for every observation
biomass_observations <- merged_fish_targets %>%
  mutate(Biomass_Species_Transect_Size = Count * (LWRa * ((LWRconv * Size)^LWRb)))
```

```{r fish4, fig.height=8, fig.width=14}
# Calculate transect-level biomass for all transect/category combinations, as sum of observation-level biomass for each transect/category combination
biomass_transect <- biomass_observations %>%
  group_by(Transect, `Length Surveyed (m)`, Biomass_Category, Code, Name.x, Subregion) %>%
  summarize(Total_Biomass_Transect_Category = sum(Biomass_Species_Transect_Size)) %>%
  full_join(complete_grid_fish, by = c("Transect", "Biomass_Category")) %>%
  mutate(Total_Biomass_Transect_Category = replace_na(Total_Biomass_Transect_Category, 0)) %>% # Ensure NAs are read as 0s
  left_join(df_fish_transect, by = "Transect") %>%
  left_join(df_metadata_survey, by = "Survey")
```

```{r fish5, fig.height=8, fig.width=14}
# Calculate transect-level biomass density (g/100m2) as transect-level biomass per m2, times 100
biomass_density_calculations <- biomass_transect %>%
  mutate(Transect_Biomass_Density = 100 * Total_Biomass_Transect_Category / (2 * `Length Surveyed (m).y`))
```

```{r fish6, fig.height=8, fig.width=14}
# Calculate biomass density (g/100m2) average by site as mean of transect biomass densities
biomass_density_average <- biomass_density_calculations %>%
  group_by(Biomass_Category, Code.y, Name, Subregion.y, `Fish Transects`) %>%
  summarise(Biomass_Density_Site_Avg = sum(Transect_Biomass_Density)) %>%
  summarise(Biomass_Density_Site_Avg = Biomass_Density_Site_Avg / `Fish Transects`)
```

```{r fish7, fig.height=8, fig.width=14}
# Calculate biomass density (g/100m2) average by subregion as mean of biomass density average by site for each subregion
biomass_density_average_subregion <- biomass_density_average %>%
  group_by(Biomass_Category, Subregion.y) %>%
  summarise(Biomass_Density_Subregion_Avg = mean(Biomass_Density_Site_Avg))
```

### 5 - Comparing the Results

The following script provides a way to compare these results to HRI official results. To change subregion, adjust the variable subregion_of_interest.

```{r fishcompare, fig.height=8, fig.width=14}
# Load in RHI 2024 calculated values
HRI_RHI <- read_excel("RHI 2024 Redo calculations with new Turneffe for reviewApril 9 2024.xlsx",
  sheet = 1, trim_ws = TRUE, na = ""
)

# Select which subregion you will compare to the HRI results
subregion_of_interest <- "Central Barrier Complex"

# Restructure the data to be comparable
HRI_RHI_subregion_herb <- HRI_RHI %>%
  filter(Subregion == subregion_of_interest) %>%
  mutate(Biomass_HRI = `Herbivorous Fish (g/100m2)`) %>%
  select(Code, Name, Biomass_HRI)
HRI_RHI_subregion_comm <- HRI_RHI %>%
  filter(Subregion == subregion_of_interest) %>%
  mutate(Biomass_HRI = `Commercial Fish (g/100m2)`) %>%
  select(Code, Name, Biomass_HRI)
calculated_RHI_subregion_herb <- biomass_density_average %>%
  filter(Subregion.y == subregion_of_interest & Biomass_Category == "H") %>%
  rename(Code = Code.y)
calculated_RHI_subregion_comm <- biomass_density_average %>%
  filter(Subregion.y == subregion_of_interest & Biomass_Category == "C") %>%
  rename(Code = Code.y)
```


```{r fishcompare_herb, fig.height=8, fig.width=14}
# Compare herbivorous
compare_fish_herb <- HRI_RHI_subregion_herb %>%
  left_join(calculated_RHI_subregion_herb, by = c("Code", "Name")) %>%
  filter(!is.na(Biomass_Density_Site_Avg)) %>%
  select(Code, Name, Biomass_HRI, Biomass_Density_Site_Avg) %>%
  mutate(Biomass_HRI = as.numeric(Biomass_HRI), Biomass_Density_Site_Avg = as.numeric(Biomass_Density_Site_Avg)) %>%
  mutate(Biomass_HRI = round(Biomass_HRI, 1), Biomass_Density_Site_Avg = round(Biomass_Density_Site_Avg, 1)) %>%
  mutate(Match = case_when(Biomass_HRI == Biomass_Density_Site_Avg ~ "Yes", Biomass_HRI != Biomass_Density_Site_Avg ~ "No"))

compare_fish_herb
```

```{r fishcompare_comm, fig.height=8, fig.width=14}
# Compare commercial
compare_fish_comm <- HRI_RHI_subregion_comm %>%
  left_join(calculated_RHI_subregion_comm, by = c("Code", "Name")) %>%
  filter(!is.na(Biomass_Density_Site_Avg)) %>%
  select(Code, Name, Biomass_HRI, Biomass_Density_Site_Avg) %>%
  mutate(Biomass_HRI = as.numeric(Biomass_HRI), Biomass_Density_Site_Avg = as.numeric(Biomass_Density_Site_Avg)) %>%
  mutate(Biomass_HRI = round(Biomass_HRI, 1), Biomass_Density_Site_Avg = round(Biomass_Density_Site_Avg, 1)) %>%
  mutate(Match = case_when(Biomass_HRI == Biomass_Density_Site_Avg ~ "Yes", Biomass_HRI != Biomass_Density_Site_Avg ~ "No"))

compare_fish_comm
```

